In the Crucible: Massive Compact Galaxies with High-Velocity Outflows and Eddington-Limited Star Formation

### 2012-04-20 trunk/tex/sigmasfr/ms.tex
first full draft

\documentclass[apj]{emulateapj}
%\documentclass[12pt,preprint]{aastex}

\newcommand{\kms}{km~s$^{-1}$}
\newcommand{\mgii}{\textrm{Mg}~\textsc{ii}}
\newcommand{\nev}{[\textrm{Ne}~\textsc{v}]}
\newcommand{\oii}{[\textrm{O}~\textsc{ii}]}
\newcommand{\oiii}{[\textrm{O}~\textsc{iii}]}
\newcommand{\msun}{M$_{\odot}$}
\newcommand{\mstar}{M$_{*}$}
\newcommand{\units}{M$_{\odot}$~yr$^{-1}$~kpc$^{-2}$}
\newcommand{\lrest}{\lambda_{\textnormal{\scriptsize{rest}}}}
\newcommand{\lobs}{\lambda_{\textnormal{\scriptsize{obs}}}}
\newcommand{\sigmasfr}{\Sigma_{\textnormal{\scriptsize{SFR}}}}
\newcommand{\sfrir}{\textnormal{SFR}_{\textnormal{\scriptsize{IR}}}}
\newcommand {\apgt} {\ {\raise-.5ex\hbox{$\buildrel>\over\sim$}}\ }
\newcommand {\aplt} {\ {\raise-.5ex\hbox{$\buildrel<\over\sim$}}\ }

\shorttitle{Compact Starbursts Revealed by HST and WISE}
\shortauthors{Diamond-Stanic et al.}

@@

%\title{HST and WISE Reveal Compact Starbursts Driving 1000
%  \kms\ Outflows}

@@

Based on infrared photometry from the Wide-field Infrared Survey
Explorer (WISE) and optical morphologies from the Hubble Space
Telescope (HST), we present the discovery of compact, obscured star
formation in galaxies that exhibit $\gtrsim1000$~\kms\ outflows.  The
implied star formation rate (SFR) surface densities approach
$\sigmasfr\approx5000$~\msun~yr$^{-1}$~kpc$^{-2}$, comparable to the
Eddington limit for a starburst where radiation pressure on dust
grains overcomes gravity.  We argue that feedback associated with this
starburst event, in the form of radiation pressure and supernova ram
pressure, is a likely mechanism for launching the high-velocity
outflows.

@@

% The relative importance of momentum injection versus thermal heating
% may vary as a function of galaxy mass \citep[e.g.,][]{hop12}, with
% thermal heating being less efficient in galaxies with larger gas
% densities and shorter cooling times.  Massive star clusters with large
% gas surface densities are a potential launching point for
% galactic-scale outflows driven by radiation pressure
% \citep[e.g.,][]{mur11}.

Galactic winds are ubiquitous in star-forming galaxies at all
redshifts and generally exhibit outflow velocities in the
100--500~\kms\ range
\citep[e.g.,][]{hec00,sha03,mar05,rup05,wei09,rub10,ste10}.  Thus, the
discovery of $|v|>1000$~\kms\ outflows in a sample of massive
($\textnormal{\mstar}\approx10^{11}$~\msun) post-starburst galaxies at
$z\sim0.6$ by \citet{tre07} suggested that a more energetic source,
such as feedback from an accreting supermassive black hole
\citep[e.g.,][]{sil98,dim05,deb12}, could have been responsible for
launching the winds.

@@

% $|v|\gtrsim1000$~\kms\

In this Letter, we present results for a sample of massive galaxies at
$z\sim0.6$ that exhibit $|v|\apgt1000$~\kms\ outflows, expanding on
the sample from \citet{tre07}.  We seek to test whether the energetic
outflows in these galaxies could have been driven by feedback from
starbursts with very large star-formation rate (SFR) surface densities
($\sigmasfr$).  Our analysis combines galaxy sizes obtained with the
Hubble Space Telescope (HST) with star-formation rates and stellar
masses estimated from WISE, Spitzer, SDSS, and GALEX photometry.

% For example, there is evidence for a positive correlation between
% outflow velocity and starburst luminosity \citep[e.g.,][]{mar05},
% albeit with significant scatter.  Furthermore, \citet{hec11} recently
% found outflows reaching velocities of $-1500$~\kms\ in a sample of
% local starbursts with compact nuclei, and argued their higher
% velocities could be explained by feedback exerted by a large number of
% massive stars that are confined to a small radius.

% [which has been invoked theoretically as
%  a mechanism to quench star formation in galaxies,

% Winds are also common in active galactic nuclei (AGNs) with velocities
% in the 1000--20,000~\kms range \citep[e.g., ][]{cre03}, although it is
% unclear whether AGN feedback plays an important role in regulating
% star formation in galaxies.  There is clear observational evidence
% that active galactic nuclei (AGNs) produce outflows
% \citep[e.g.,][]{cre03} and that the outflowing gas can extend to
% $>$kpc scales \citep[e.g.,][]{rup11}, it is unclear whether such
% feedback plays an important role globally in quenching star formation.

% Feedback from massive stars and active black holes has important
% implications for galaxy evolution (e.g., luminosity function, star
% formation, galaxy color bi-modality, mass-metallicity relation,
% enrichment of IGM).

% Feedback from accreting supermassive black holes has also been invoked
% theoretically as a mechanism for quenching star formation in galaxies
% \citep[e.g.,][]{sil98,gra04,dim05}.  There is clear observational
% evidence that active galactic nuclei (AGNs) produce outflows
% \citep[e.g.,][]{cre03} and that the outflowing gas can extend to
% $>$kpc scales \citep[e.g.,][]{rup11}, but it is unclear whether such
% feedback plays an important role globally in quenching star formation
% and shaping the galaxy mass function.

% and optical spectra ($\lrest=2500$-5200~\AA)

%  The spectral region surrounding the
%  \mgii~$\lambda\lambda2796,2803$ absorption alines, which are used to
%  measure outflow velocities, is highlighted in both wavelength and
%  velocity space.

%  and are scaled to match the WISE channel 2 photometry ($\lobs=4.6$~$\mu$m)
% (\mgii~$\lambda2796$ is plotted in blue and \mgii~$\lambda2803$ is plotted in red)

@@

%\subsection{HST/WFC3 Imaging}
\subsection{Morphologies and Sizes}

@@

\begin{figure}[!t]
%\includegraphics[angle=90,scale=0.68]{sigmasfr_datafigure.ps}
\includegraphics[angle=0,scale=0.41]{3seds.ps}
\caption{Left: Observed UV--IR SEDs ($\lrest=0.1$--15~$\mu$m) for
  three galaxies with large SFR surface densities
  ($\sigmasfr>3000$~\units).  We show stellar population fits to the
  $\lrest=0.1$--3~$\mu$m emission (black solid line) and three
  templates for dust emission \citep[M82 starburst, Arp 220 starburst,
    and obscured quasar;][]{pol07}.  The starburst templates provide
  reasonable fits to the $\lobs=12$ and 22~$\mu$m WISE photometry,
  while the quasar template does not.  Right: HST/WFC3 F814W images
  (probing $\lrest\approx5000$~\AA) showing that these galaxies are
  dominated by a compact nucleus.}
\label{fig:seds}
\end{figure}
% scaled to the WISE $\lobs=4.6~\mu$m flux.

@@

%\subsection{Photometry, Star Formation Rates, and Stellar Masses}
\subsection{Star Formation Rates and Stellar Masses}

@@

\begin{figure*}[!t]
\begin{center}
\includegraphics[angle=0,scale=.8]{sigmasfr.ps} %\\
%\includegraphics[angle=0,scale=.7]{sizemass_sigmasfr.ps}
%\caption{$\sigmasfr$--mass and size--mass relationships.}
\caption{SFR surface densities and stellar masses for the HST--WISE
  sample described in this paper (black circles, with symbol size
  proportional to outflow velocity), along with samples of gas-rich
  mergers (orange squares) and star-forming galaxies (shown with 68\%,
  95\%, and 99.7\% contours).  We mark the empirical threshold for
  launching winds \citep[dotted line,
    $\sigmasfr\approx0.1$~\units;][]{hec02}, the 90th-percentile
  starburst intensity limit from \citet{meu97} (dashed line,
  $\sigmasfr\approx45$~\units), and the Eddington limit from radiation
  pressure on dust grains \citep[solid line,
    $\sigmasfr\approx2000$~\units;][]{mur05,tho05,hop10}.  Our
  HST--WISE sample overlaps with the region characterized by gas-rich
  mergers, and extends to very large SFR surface densities near the
  Eddington limit, suggesting growth that is limited by momentum
  injection from massive stars.}
\label{fig:sigmasfr}
\end{center}
\end{figure*}

% The sample of gas-rich mergers includes ULIRGs
%   \citep{vei06}, Lyman break analogs with dominant central objects
%   \citep{ove09}, and the local compact starburst Arp 220 \citep{sco97,
%     ken98, rod08}.  The samples of star-forming galaxies are drawn
%   from \citet{wuy11} at $0.02<z<0.2$ (solid black contour),
%   $0.5<z<1.5$ (dashed red countour), and $1.5<z<2.5$ (dotted purple
%   contour), and from \citet{law12} at $2.5<z<3.6$ (dash-dotted blue
%   contour).

%\subsection{Optical Spectroscopy}
\subsection{Outflow Kinematics and Covering Factors}

@@

% MMT/Blue Channel (J1506+5402, $R\sim1000$), SDSS (J1506+5402,
% $R\sim2000$), Magellan/MAGE (J1341-0321, $R\sim4000$), Keck/LRIS
% (J0905+5759, $R\sim600$), and Keck/HIRES (J0905+5759, $R\sim37,000$)

%\section{Analysis}
%
%\subsection{Star-formation Rates and Stellar Masses}
%
%\subsection{Morphologies and Sizes}
%
%\subsection{Outflow Kinematics and Covering Factors}

% \begin{figure*}[!t]
% \begin{center}
% \includegraphics[angle=0,scale=.8]{sizemass_sigmasfr.ps}
% \caption{size--mass relationship.}
% \label{fig:sizemass}
% \end{center}
% \end{figure*}

% %\begin{center}
% \begin{deluxetable}{ccccccc}
% \tabletypesize{\scriptsize}
% 
% \tablecaption{Galaxy Properties\label{tab:properties}}
% \tablewidth{0pt}
% 
% \tablehead{ \colhead{Galaxy Name} & \colhead{z} & \colhead{$M_{*}$} & \colhead{$v$} & \colhead{SFR} & \colhead{$r_e$} & \colhead{$\sigmasfr$} }
% 
% \startdata
% J150636.29+540220.7   & 0.608 & 11.01 & -1211 &  621 & 0.13 & 5851 \\
% J090523.60+575912.4   & 0.712 & 10.62 & -2470 & 344 & 0.10 & 5478 \\ 
% %J094417.84+093019.3   & 0.514 & 10.73 & -1778 & $<43.22$ & 161 & 0.15 & 1139 \\
% J134136.79$-$032125.2   & 0.658 & 10.61 & -875  & 490 & 0.16 & 3058 \\ 
% % J1506 has Lx=42.66
% \enddata
% 
% %\tablecomments{}
% %\tablenotetext{a}{Star-formation rate based on \citet{ken98}.  No extinction correction has been applied.}
% 
% \end{deluxetable}
% %\end{center}

@@

The compact sizes ($r_e\approx100$~pc) and large star formation rates
($\textnormal{SFR}\approx500$~\msun) for the galaxies described above
implies extremely large star-formation rate surface densities
($\sigmasfr\approx5000$~\units).  To place these galaxies in context,
we plot their $\sigmasfr$ values as a function of stellar mass in
Figure~\ref{fig:sigmasfr}.  We include comparison samples of
$\sim10^5$ star-forming galaxies at $0.5<z<1.5$ from \citet{wuy11} and
a few dozen gas-rich mergers at $z<0.3$ including 32 ULIRGs from
\citet{vei06}, five Lyman break analogs with dominant central objects
from \citet{ove09}, and the local compact starburst Arp 220
\citep{sco97, ken98, rod08}.  We also mark the empirical threshold for
launching winds \citep[$\sigmasfr\approx0.1$~\units,][]{hec02}, the
90th percentile limit for the surface brightness of starbursts
measured using UV, H$\alpha$, far-IR, and radio continuum emission
\citep[$\sigmasfr\approx45$~\units][]{meu97}, and the theoretical
limit for a starburst limited by feedback from radiation pressure
\citep[$\sigmasfr\approx2000$~\units,][]{mur05,tho05,hop10}.  The most
luminous, compact starbursts in our sample exhibit SFR surface
densities that reach the Eddington limit, suggesting that their growth
is being regulated by momentum input from massive stars.

%  We also include measurements of
% late-stage, gas-rich mergers at $z<0.3$ from \citet{vei06},
% star-forming galaxies at $z<0.3$ from \citet{ove09}, and star-forming
% galaxies at $z=1.5$--3.6 from \citet{law12}.

% Points to make:
% \begin{itemize}
% \item \citet{ken98}, the standard reference for the relationship
%   between SFR and gas surface density, includes Arp 200 with
%   $\sigmasfr\approx10^3$~\units\ and
%   $\Sigma_{H_2}\approx5\times10^{10}$~\msun~kpc$^{-2}$.  
% \item The star clusters considered by \citet{meu97} extend into this
%   $\sigmasfr>10^{3}$~\units\ range.
% 
% \end{itemize}

% We show relationship between effective radius and stellar mass for our
% WISE-HST sample in bottom panel of Figure~\ref{fig:sizemass}.  Here,
% we include the gas-rich mergers from \citet{vei06} as well as
% $z\sim2.3$ compact massive galaxies from \citet{van08}.  We also show
% the size--mass relationship for elliptical galaxies at $z\sim0.1$ from
% \citet{she03}, and the average size evolution of spheroidal ($n>2.5$)
% galaxies from $z=2$ to $z=0.1$ from \citet{tru07}.

%\subsection{A Maximum Star-Formation Rate Surface Density?}

%\section{Discussion}

@@

% ased on its \oiii\ luminosity $L(\oiii)=10^{42.1}$~erg~s$^{-1}$, the
% expected X-ray luminosity \citep[adopting the relationship between
%   \oiii\ and X-ray luminosity for type 1 AGNs from][]{hec05}

% No evidence in spectra for unobscured AGN component.  J1506+5402 has
% X-ray and \nev.  Agreement between SFR indicators.  Mention Balmer
% lines and dust geometry.  SEDs well fit by star-forming tempates.
% Most ULIRGs are starburst dominated.  Mention Pauls's paper.  There is
% likely some ongoing AGN activity in sources like J1506+5402, but in
% general, the AGN does not seem to be relevant energetically.  Point
% that high-velocity outflows don't mean that an outflow is AGN driven.

\subsection{The Outflow Launching Mechanism}

@@

% Following \citet{hec93} and
% \citet{hec11}, the expected terminal velocity for a cloud with column
% density $\approx10^{21}$~cm$^{-2}$ launched from a radius
% $\approx200$~pc and driven by momentum flux
% $\approx2\times10^{35}$~dynes is:
% \begin{eqnarray}\nonumber
% v_{\infty}&=&1800~\textnormal{km~s}^{-1}
% \left(\dot{p}\over2\times10^{35}~\textnormal{dyne}\right)^{1/2} \\
% &&
% \left({r\over200~\textnormal{pc}}\right)^{-1/2}
% \left({N\over10^{21}~\textnormal{cm}^{-2}}\right)^{-1/2}
% \label{eq:vterm}
% \end{eqnarray}
% The expected momentum flux from stellar winds and supernovae for a
% starburst with $\textnormal{SFR}\approx200$~\msun~yr$^{-1}$ is
% $\approx5$--$10\times10^{35}$~dynes \citep{lei92,hec93,lei99,vei05}
% while the $L_{bol}/c$ radiation pressure is
% $\approx2\times10^{35}$~dynes.  Thus, there is more than enough
% momentum available from stellar winds and supernovae or from radiation
% pressure to produce the outflow velocities we observe.

%AGN have been invoked to launch high-velocity winds.

% \subsection{Momentum Flux from Supernovae and Stellar Winds}
% 
% \citet{meu97} argued for a limit near 45~\units (somewhat smaller for
% a Chabrier IMF) based on the surface brightness of starbursts measured
% using rest-frame UV, H$\alpha$, far-IR, and radio continuum emission,
% such that only 10\% of starbursts exceeded this limit.  They argued
% that the momentum flux from supernovae and stellar winds expected at
% this threshold matches the mean central pressure measured by
% \citet{hec90} for starburst-driven winds.
% 
% There are also relevant arguments about ram pressure driving of
% high-velocity winds from \citet{hec11}.
% 
% 
% \subsection{The Eddington Limit from Radiation Pressure on Dust Grains}

% In models where gaseous outflows are driven by radiation pressure from
% massive stars, estimates of the maximum $\sigmasfr$ where the outward
% force from radiation exceeds the inward force of gravity are in the
% $\sigmasfr\sim10^3$~\units\ range \citep[e.g.,][]{tho05,hop10}.

% Our galaxies exceed the \citet{meu97} limit but are roughly in line
% with the limit from radiation pressure.

% The relative importance of momentum injection versus thermal heating
% may vary as a function of galaxy mass \citep[e.g.,][]{hop12}, with
% thermal heating being less efficient in galaxies with larger gas
% densities and shorter cooling times.  Regarding momentum injection,
% \citet{sha11} have argued that radiation pressure becomes more
% important than ram pressure at large SFRs.

%\subsection{Outflows at the Escape Velocity}

% This is in agreement with the predicitons from radiation pressure
% launching winds from the most massive star clusters in a galaxy.

% Question of whether other observables are plausible.  
% \begin{itemize}
% \item Expected velocity dispersion.  Some results have suggested that
%   turbulence becomes large at high surface densities, and therefore
%   the velocity dispersion should be large.  
% \item The dynamical time for the starburst.  If gas is consumed with
%   an efficiency of a few \% per dynamical time, how long do we expect
%   the starburst to last?
% \end{itemize}
% 
% Other points: 
% \begin{itemize}
% \item Is this mechanism viable for the whole sample?  We see large
%   outflow velocities in rest of the sample, where the galaxies are
%   more ``post'' starburst.  This fits in a scenario where the genuine
%   post-starbursts are further along in the evolution after the outflow
%   quenches star formation.
% \item Will these evolve into ellipticals, or do they have enough gas
%   to reform a disk?  If they do quench, how difficult would it be to
%   grow in size to match the local size--mass relationship?  Could
%   their descendants be the handful of local galaxies that don't lie on
%   the size--mass relationship?
% \item Is there disagreement between $L_{IR}+L_{UV}$ and H$\alpha$
%   SFRs?  \citet{hec11} pointed out that this could be a sign of
%   escaping ionizing photons.
% \end{itemize}

@@

They are rare and thus represent an unusual or short-lived phase.  Gas
consumption time for starbursts.  Connection to ellipticals.
Conceivable that they could grown in size sufficiently by $z=0$?

Few galaxies exhibit such high-velocity outflows because few galaxies
have such extreme SFR surface densities.

Connection to models.  How such large SFRs given the powerful wind?

High-velocity outflows do not imply AGN feedback.

Theoretical arguments about ram v. radiation pressure.  Point that we
are observing 10$^4$~K gas.

Sample as a whole. Interesting that high-$\sigmasfr$ galaxies tend to
have larger outflow velocities.

Mention high-$\sigmasfr$ galaxies at high z.

@@

@@

% James Bullock, Xavier Prochaska, Kate Rubin, Art Wolfe, Alex Mendez

@@

% \bibitem[Baldwin et al.(1981)]{bal81} Baldwin, J.~A., Phillips, M.~M.,
%   \& Terlevich, R.\ 1981, \pasp, 93, 5
% \bibitem[Barton \& Cooke(2009)]{bar09} Barton, E.~J., \& Cooke,
%   J.\ 2009, \aj, 138, 1817
% \bibitem[Bordoloi et al.(2011)]{bor11} Bordoloi, R., et al.\ 2011,
%   arXiv:1106.0616
% \bibitem[Buitrago et al.(2008)]{bui08} Buitrago, F., Trujillo, I.,
%   Conselice, C.~J., Bouwens, R.~J., Dickinson, M., \& Yan, H.\ 2008,
%   \apjl, 687, L61
% \bibitem[Coil et al.(2011)]{coi11} Coil, A.~L., Weiner, B.~J., Holz,
%   D.~E., et al.\ 2011, \apj, 743, 46

@@

% \bibitem[Chen et al.(2010)]{hchen10} Chen, H.-W., Wild, V., Tinker,
%   J.~L., Gauthier, J.-R., Helsby, J.~E., Shectman, S.~A., \& Thompson,
%   I.~B.\ 2010, \apjl, 724, L176
% \bibitem[Chen et al.(2010b)]{ychen10} Chen, Y.-M., Tremonti, C.~A.,
%   Heckman, T.~M., Kauffmann, G., Weiner, B.~J., Brinchmann, J., \&
%   Wang, J.\ 2010, \aj, 140, 445

@@

% \bibitem[Coil et al.(2011)]{coi11} Coil, A.~L., Weiner, B.~J., Holz,
%   D.~E., Cooper, M.~C., Yan, R., \& Aird, J.\ 2011, arXiv:1104.0681
\bibitem[Crenshaw et al.(2003)]{cre03} Crenshaw, D.~M., Kraemer,
  S.~B., \& George, I.~M.\ 2003, \araa, 41, 117
% \bibitem[Croton et al.(2006)]{cro06} Croton, D.~J., et al.\ 2006,
%   \mnras, 365, 11

@@

% \bibitem[Dekel \& Birnboim(2006)]{dek06} Dekel, A., \& Birnboim,
%   Y.\ 2006, \mnras, 368, 2

@@

\bibitem[Granato et al.(2004)]{gra04} Granato, G.~L., De Zotti, G.,
  Silva, L., Bressan, A., \& Danese, L.\ 2004, \apj, 600, 580
\bibitem[Heckman et al.(1990)]{hec90} Heckman, T.~M., Armus, L., \&
  Miley, G.~K.\ 1990, \apjs, 74, 833

@@

% \bibitem[Hopkins et al.(2009)]{hop09} Hopkins, P.~F., Bundy, K.,
%   Murray, N., Quataert, E., Lauer, T.~R., \& Ma, C.-P.\ 2009, \mnras,
%   398, 898
%\bibitem[Hopkins et al.(2009)]{hop09} Hopkins, P.~F., Hernquist, L.,
%  Cox, T.~J., Keres, D., \& Wuyts, S.\ 2009, \apj, 691, 1424

@@

% \bibitem[Hopkins et al.(2010a)]{hop10a} Hopkins, P.~F.,
%   Bundy, K., Hernquist, L., Wuyts, S., \& Cox, T.~J.\ 2010, \mnras,
%   401, 1099

@@

% %\bibitem[Hopkins et al.(2011)]{hop11} Hopkins, P.~F., Quataert, E., \&
% %  Murray, N.\ 2011, arXiv:1101.4940
% \bibitem[Hopkins et al.(2011)]{hop11} Hopkins, P.~F., Quataert, E., \&
%   Murray, N.\ 2011, arXiv:1110.4638
\bibitem[Hopkins et al.(2012)]{hop12} Hopkins, P.~F., Quataert, E., \&
  Murray, N.\ 2012, \mnras, arXiv:1110.4638
% \bibitem[Kacprzak et al.(2010)]{kac10} Kacprzak, G.~G., Churchill,
%   C.~W., Ceverino, D., Steidel, C.~C., Klypin, A., \& Murphy,
%   M.~T.\ 2010, \apj, 711, 533
% \bibitem[Kacprzak et al.(2011a)]{kac11a} Kacprzak, G.~G., Churchill,
%   C.~W., Barton, E.~J., \& Cooke, J.\ 2011, \apj, 733, 105
% \bibitem[Kacprzak et al.(2011b)]{kac11b} Kacprzak, G.~G., Churchill,
%   C.~W., Evans, J.~L., Murphy, M.~T., \& Steidel, C.~C.\ 2011,
%   arXiv:1106.3068
% \bibitem[Kauffmann et al.(2003)]{kau03} Kauffmann, G., Heckman, T.~M.,
%   Tremonti, C., et al.\ 2003, \mnras, 346, 1055
% \bibitem[Kennicutt(1998)]{ken98} Kennicutt, R.~C., Jr.\ 1998, \araa,
%   36, 189

@@

% \bibitem[Kere{\v s} et al.(2005)]{ker05} Kere{\v s}, D., Katz, N.,
%   Weinberg, D.~H., \& Dav{\'e}, R.\ 2005, \mnras, 363, 2
% \bibitem[Kere{\v s} et al.(2009)]{ker09} Kere{\v s}, D., Katz, N.,
%   Dav{\'e}, R., Fardal, M., \& Weinberg, D.~H.\ 2009, \mnras, 396,
%   2332
% \bibitem[Kobulnicky \& Kewley(2004)]{kob04} Kobulnicky, H.~A., \&
%   Kewley, L.~J.\ 2004, \apj, 617, 240

@@

% \bibitem[Kriek et al.(2010)]{kri10} Kriek, M., et al.\ 2010, \apjl,
%   722, L64
\bibitem[Law et al.(2012)]{law12} Law, D.~R., Steidel, C.~C., Shapley,
  A.~E., et al.\ 2012, \apj, 745, 85

@@

% \bibitem[Lehnert \& Heckman(1996)]{leh96} Lehnert, M.~D., \& Heckman,
%   T.~M.\ 1996, \apj, 462, 651

@@

@@

% \bibitem[Martin(2006)]{mar06} Martin, C.~L.\ 2006, \apj, 647, 222

>>>>

\documentclass[apj]{emulateapj}

\newcommand{\kms}{km~s$^{-1}$}
\newcommand{\mgii}{\textrm{Mg}~\textsc{ii}}
\newcommand{\nev}{[\textrm{Ne}~\textsc{v}]}
\newcommand{\oii}{[\textrm{O}~\textsc{ii}]}
\newcommand{\oiii}{[\textrm{O}~\textsc{iii}]}
\newcommand{\msun}{M$_{\odot}$}
\newcommand{\mstar}{M$_{*}$}
\newcommand{\units}{M$_{\odot}$~yr$^{-1}$~kpc$^{-2}$}
\newcommand{\lrest}{\lambda_{\textnormal{\scriptsize{rest}}}}
\newcommand{\lobs}{\lambda_{\textnormal{\scriptsize{obs}}}}
\newcommand{\sigmasfr}{\Sigma_{\textnormal{\scriptsize{SFR}}}}
\newcommand{\sfrir}{\textnormal{SFR}_{\textnormal{\scriptsize{IR}}}}

\shorttitle{Compact Starbursts Driving High-Velocity Outflows}
\shortauthors{Diamond-Stanic et al.}

@@

@@

We present the discovery of compact, obscured star formation in
galaxies that exhibit $\gtrsim1000$~\kms\ outflows.  Using optical
morphologies from the Hubble Space Telescope and infrared photometry
from the Wide-field Infrared Survey Explorer, we estimate star
formation rate (SFR) surface densities that approach
$\sigmasfr\approx5000$~\msun~yr$^{-1}$~kpc$^{-2}$, which is comparable
to the Eddington limit from radiation pressure on dust grains.  We
argue that feedback associated with a compact starburst in the form of
radiation pressure from massive stars and ram pressure from supernovae
and stellar winds is sufficient to produce the high-velocity outflows
we observe, without the need to invoke feedback from an active
galactic nucleus.

@@

Galactic winds are ubiquitous in star-forming galaxies at all
redshifts and generally exhibit outflow velocities in the
100--500~\kms\ range
\citep[e.g.,][]{hec00,sha03,mar05,rup05,wei09,rub10,ste10}.  Thus, the
discovery of $|v|>1000$~\kms\ outflows in a sample of massive
($\textnormal{\mstar}\approx10^{11}$~\msun) post-starburst galaxies at
$z\sim0.6$ by \citet{tre07} suggested that a more energetic source,
such as feedback from an accreting supermassive black hole
\citep[e.g.,][]{sil98,dim05,deb12}, could have been responsible for
launching the winds \citep[see][for a recent review]{fab12}.

@@

In this Letter, we present results for a sample of massive galaxies at
$z\sim0.6$ that exhibit $\gtrsim1000$~\kms\ outflows, expanding on the
sample from \citet{tre07}.  We seek to test whether the energetic
outflows in these galaxies could have been driven by feedback from
starbursts with very large star-formation rate (SFR) surface densities
($\sigmasfr$).  Our analysis combines galaxy sizes obtained with the
Hubble Space Telescope (HST) with star-formation rates and stellar
masses estimated from WISE, Spitzer, SDSS, and GALEX photometry.

\begin{figure}[!t]
\includegraphics[angle=0,scale=0.41]{3seds.ps}
\caption{Left: Observed UV--IR SEDs ($\lrest=0.1$--15~$\mu$m) for
  three galaxies with large SFR surface densities
  ($\sigmasfr>3000$~\units).  We show stellar population fits to the
  $\lrest=0.1$--3~$\mu$m emission (black solid line) and three
  templates for dust emission \citep[M82 starburst, Arp 220 starburst,
    and obscured quasar;][]{pol07}.  The starburst templates provide
  reasonable fits to the $\lobs=12$ and 22~$\mu$m WISE photometry,
  while the quasar template does not.  Right: HST/WFC3 F814W images
  (probing $\lrest\approx5000$~\AA) showing that these galaxies are
  dominated by a compact nucleus.}
\label{fig:seds}
\end{figure}

@@

\subsection{Morphologies and Sizes}

@@

@@

\subsection{Star Formation Rates and Stellar Masses}

@@

\begin{figure*}[!t]
\begin{center}
\includegraphics[angle=0,scale=.8]{sigmasfr.ps} 
\caption{SFR surface densities and stellar masses for the HST--WISE
  sample described in this paper (black circles, with symbol size
  proportional to outflow velocity), along with samples of gas-rich
  mergers (orange squares) and star-forming galaxies (shown with 68\%,
  95\%, and 99.7\% contours).  We mark the empirical threshold for
  launching winds \citep[dotted line,
    $\sigmasfr\approx0.1$~\units;][]{hec02}, the 90th-percentile
  starburst intensity limit from \citet{meu97} (dashed line,
  $\sigmasfr\approx45$~\units), and the Eddington limit from radiation
  pressure on dust grains \citep[solid line,
    $\sigmasfr\approx2000$~\units;][]{mur05,tho05,hop10}.  Our
  HST--WISE sample overlaps with the region characterized by gas-rich
  mergers, and extends to very large SFR surface densities near the
  Eddington limit, suggesting growth that is limited by momentum
  injection from massive stars.}
\label{fig:sigmasfr}
\end{center}
\end{figure*}

\subsection{Outflow Kinematics and Covering Factors}

@@

@@

The compact sizes ($r_e\approx100$~pc) and large star formation rates
($\textnormal{SFR}\approx300$~\msun) for the galaxies described above
implies extremely large star-formation rate surface densities
($\sigmasfr\approx5000$~\units).  To place these galaxies in context,
we plot their $\sigmasfr$ values as a function of stellar mass in
Figure~\ref{fig:sigmasfr}.  We include comparison samples of
$\sim10^5$ star-forming galaxies at $0.5<z<1.5$ from \citet{wuy11} and
a few dozen gas-rich mergers at $z<0.3$ including 32 ULIRGs from
\citet{vei06}, five Lyman break analogs with dominant central objects
from \citet{ove09}, and the local compact starburst Arp 220
\citep{sco97, ken98, rod08}.  We also mark the empirical threshold for
launching winds \citep[$\sigmasfr\approx0.1$~\units,][]{hec02}, the
90th percentile limit for the surface brightness of starbursts
measured using UV, H$\alpha$, far-IR, and radio continuum emission
\citep[$\sigmasfr\approx45$~\units][]{meu97}, and the theoretical
limit for a starburst limited by feedback from radiation pressure
\citep[$\sigmasfr\approx2000$~\units,][]{mur05,tho05,hop10}.  The most
luminous, compact starbursts in our sample exhibit SFR surface
densities that reach the Eddington limit, suggesting that their growth
is being regulated by momentum input from massive stars.

@@

\subsection{The Outflow Launching Mechanism}\label{sec:launch}

@@

@@

It is clear from Figure~\ref{fig:sigmasfr} that the high-$\sigmasfr$
galaxies in our sample constitute a rare population, suggesting that
they represent an unusual or short-lived phase.  Mergers of gas-rich
galaxies are a viable mechanism for producing compact starbursts
\citep[e.g.,][]{mih96}, and such gas-rich major mergers are rare at
$z\sim0.6$ due to the decline in both the merger rate
\citep[e.g.,][]{lot11} and the gas fraction \citep[e.g.,][]{tac10} of
galaxies since $z\sim2$.  Furthermore, the high-$\sigmasfr$ galaxies
are caught in a particular time interval where there is both vigorous
star formation and strong feedback.  The length of this phase may be
set by the gas consumption timescale or the timescale for feedback to
suppress subsequent star formation.  Based on the Kennicutt--Schmidt
(K-S) relation \citep{ken98}, a compact starburst with
$r_e\approx100$~pc, $\textnormal{SFR}\approx300$~\msun, and
$\sigmasfr\approx5000$~\units\ would have a gas surface density of
$\Sigma_{gas}\sim2\times10^{11}$~\msun~kpc$^{-2}$ corresponding to
$M_{gas}\sim5\times10^{9}$~\msun\ inside 100~pc, which would be
consumed on a timescale $\sim30$~Myr.  This scenario could be tested
with CO observations of molecular gas masses and kinematics for these
extreme galaxies.

Considering our full sample of galaxies in Figure~\ref{fig:sigmasfr},
a vertical $\sigmasfr$ sequence is apparent.  This could be explained
as an evolutionary sequence where the high-$\sigmasfr$ galaxies
represent the peak of the starburst where the high-velocity outflows
are launched, while the lower $\sigmasfr$ galaxies represent a
subseqent, post-starburst phase.  It is interesting to note that all
12/29 galaxies above the $\sigmasfr\approx45$~\units\ limit from
\citet{meu97} exhibit outflows (with median centroid velocity
$v=-1500$~\kms), while all 7/29 galaxies without detected outflows
have $\sigmasfr<30$~\units.  If a compact starburst is a requirement
for the production of high-velocity outflows, it may be that the
sources without outflows have not gone through such a phase (see
A. Robaina, et al., in preparation for a discussion of the
relationship between galaxy morphology, outflow velocity, and stellar
population age in this sample).

Finally, it is worthwhile to consider the implications of our results
for models of massive galaxy formation.  Simulaitons of major galaxy
mergers with large gas fractions ($f_{gas}\sim50$\%) can produce
$\textnormal{\mstar}\sim10^{11}$~\msun\ remnants with effective radii
$r_{e}\sim1$~kpc \citep[e.g.,][]{wuy10}, but our sample includes
galaxies of similar mass that are smaller by almost an order of
magnitude.  If these systems do evolve into quiescent elliptical
galaxies, it would be quite challenging for them to grow in size from
$r_e\sim0.1$~kpc to $r_e\sim5$~kpc to reach the local size--mass
relation \citep{shen03} in the $t\sim6$~Gyr since $z=0.6$.  For
comparison the compact, quiescent galaxies observed at $z\sim2$
\citep[e.g.,][]{tru07,van08} have $t\sim10$~Gyr to grow by a factor of
$\sim5$.  In this context, it is worth noting that the half-light
radius (ignoring dust attenuation) at rest-frame $V$ band (where our
HST observations probe) can be a factor of 5--10 smaller than the
half-mass radius for a gas-rich merger at final coalescence near the
peak of starburst activity \citep{wuy10}.  It would be worthwhile to
test for such size discrepancies and to pronbe the radial dependence
of the mass-to-light ratio for these galaxies by measuring sizes at
rest-frame near-IR wavelengths.

@@

We have measured large SFR surface densities for galaxies that exhibit
$\gtrsim1000$~\kms\ outflows.  The largest $\sigmasfr$ values are
comparable to the Eddington limit from radiation pressure on dust
grains, and such compact starburst are expected to have substantial
momentum input from both massive stars and supernovae.  There is some
evidence for AGN activity in a fraction of these systems, but the AGN
is generally not relevant in a bolometric sense.  High-velocity
outflows have been previously interpreted as a signpost of AGN
feedback, but we conclude that stellar feedback from a compact
starburst is equally capable of producing such a signature.

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

\bibitem[Lotz et al.(2011)]{lot11} Lotz, J.~M., Jonsson, P., Cox,
  T.~J., et al.\ 2011, \apj, 742, 103

@@
